NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

### 2022-10-12 SI.tex
Update text

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/permeationSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:overlay}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/LateralDiffusionSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:overlay}
\end{figure}

>>>>

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/permeationSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:permeationSI}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = 180mm]{Figures/LateralDiffusionSI.pdf}
    \caption{a) Structure of an overlay databank. 
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    The figures and numbers are created on 9th of May 2022.}
    \label{fig:diffusionSI}
\end{figure}

\section{Databank content}

\begin{figure}
  \includegraphics[width=\textwidth]{Figures/DataBankChart.pdf}
  \caption{Structure of the NMRlipids databank. Manually added input data (blue boxes) includes basic information on the simulation, permanent links to the raw data, and experimental data if available. The databank entries (red box) and analysis results (green boxes), locating at \url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations} are automatically generated by the computer programs included in the NMRlipids databank (yellow boxes). Because raw data are not permanently stored but can be accessed based on the information in the databank, this connection is marked with the dashed line. }\label{DatabankStructure}
\end{figure}

\begin{table}[]
    \centering
    \begin{tabular}{  p{3.5cm}  p{9.5cm}  p{4.0cm} }
    \toprule
    key & description & type  \\
    \midrule
        %\hline
    DOI & DOI from where the raw data is found & user given (compulsory) \\
    SOFTWARE & Software used to run the simulation (e.g. Gromacs, Amber, NAMD, etc.) & \\
    TRJ & Name of the trajectory file found from DOI & \\
    TPR & Name of the topology file found from DOI (trp file in the case of Gromacs) & \\
    PREEQTIME & Pre-equilibrate time simulated before the uploaded trajectory in nanoseconds. \tablefootnote{For example, if you upload 100-200 ns part of total 200 ns simulation, this should value should be 100.} & \\
    TIMELEFTOUT & Equilibration period in the uploaded trajectory that should be discarded in analyses. \tablefootnote{For example, if you upload 0-200 ns part of total 200 ns simulation where the first 100 ns should be considered as an equilibration, this value should be 100.} \\
    COMPOSITION & Molecules names used in the simulation and corresponding mapping files (see section \ref{naming}) & \\
    DIR\_WRK & Temporary working directory in your local computer. \\
    UNITEDATOM\_DICT & Information for constucting hydrogens for united atom simulations, empty for all atom simulations & \\
    TYPEOFSYSTEM & Lipid bilayer or something else & \\
    \hline
    PUBLICATION & Give reference to a publication(s) related to the data. & User given (optional)\\
    AUTHORS\_CONTACT & Name and email of the main author(s) of the data. & \\
    SYSTEM & System description on free text format & \\
    SOFTWARE\_VERSION & Version of the used software & \\
    FF & Name of the used force field & \\
    FF\_SOURCE & Source of the force field parameters, e.g, CHARMM-GUI, webpage, citation to a publication, etc. & \\
    FF\_DATE &  Date when force field parameters were accessed on the gives source (day/month/year). & \\
    FF{molename} & Molecule specific force field information, e.g., water model with FFSOL and sodium parameters with FFSOD. & \\
    CPT & Name of the Gromacs checkpoint file. & \\
    LOG & Name of the Gromacs log file. & \\
    TOP & Name of the Gromacs top file. & \\
    GRO & Name of the Gromacs gro file. & \\
    \hline
    TRAJECTORY\_SIZE & Size of the trajectory file in bytes & automatically extracted data. \\
    TRJLENGTH & Lenght of the trajectory (ps). & \\
    TEMPERATURE & Temperature of the simulation. & \\
    NUMBER\_OF\_ATOMS & Number of atoms in the simulation. & \\
    DATEOFRUNNIG & Date when added into the databank & \\
    EXPERIMENT & Potentially connected experimental data & \\
    COMPOSITION & Numbers of lipid molecules (NPOPC, NPOPG, etc.) per membrane leaflet are calculated by determining on which side of the center of mass of the membrane the center of mass of the head group of each lipid molecule is located.
    Numbers of other molecules such as solvent and ions (NSOL, NPOT, NSOD, etc.) are read from the topology file. & \\
    \end{tabular}
    \caption{Keys stored in the README.yaml files of simulations.}
    \label{tab:READMEkeys}
\end{table}

\begin{table}[h]
    \centering
    \begin{tabular}{c|c}
        Abbreviation & Molecule name \\
        \hline
        POPC &  1-palmitoyl-2-oleoyl-sn-glycero-3-phosphocholine\\
        POPG &  1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoglycerol \\
        POPS & 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-L-serine \\
        POPE & 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine \\
        CHOL & cholesterol \\
        DHMDMAB & dihexadecyldimethylammonium \\
        \hline
        POT & potassium ion \\
        SOD & sodium ion \\
        CLA & chloride ion \\
        CAL & calcium ion \\
        SOL & water \\
    \end{tabular}
    \caption{Abbreviations used in the databank}
    \label{tab:abbreviations}
\end{table}

\begin{table}[]
    \centering
    \begin{tabular}{  p{5.0cm}  p{10.0cm}}
    \toprule
    key & description \\
    \midrule
    DOI & DOI of the publication related to the experimental data. \\
    TEMPERATURE & Temperature of the experiment. \\
    MOLAR\_FRACTIONS & Dictionary of molar fractions of bilayer components \\
    ION\_CONCENTRATIONS & Dictionary of ion concentrations of the system \\% (defined as ??) \\
    TOTAL\_LIPID\_CONCENTRATION & Total concentration of lipid components. If exact concentration is not known, but experiments are performed in excess water, 'full hydration' can be given. \\
    COUNTER\_IONS & Type of counter ions if present.
\end{tabular}
    \caption{Keys stored in the README.yaml files of experiments.}
    \label{tab:READMEkeysEXP}
\end{table}

### 2022-10-12 main.tex

\begin{figure}[tb]
    \centering
    \includegraphics[width=180mm]{Figures/permeation2.pdf}
    \caption{A 
    {\bf B-E} Water permeation through membranes analyzed from the databank as a function of temperature, thickness, area per lipid, and acyl chain order. Values from simulations with non-zero permeation values are shown with blue dots. Averages over fixed range of x-axis values are shown with black dots. Insert in B) shows the Arrhenius plot of permeation ($\ln$ (P) vs. 1/T) that gives 17\,$\pm$\,4\,k$_b$T for the average activation energy for water permeation through lipid bilayer. Inserts in C) and D) show the region where the dependence could be considered approximately linear.
    {\bf F-G} Diffusion anistoropy of water as a function of thickness and area per lipid.
    {\bf H} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.}
    \label{fig:permeability}
\end{figure}

To analyze the anisotropic diffusion of water in membrane environment, we also calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank. The results at different hydration levels are shown in Fig.~\ref{fig:permeability} H together with the experimental data~\cite{rudakova04}. 
%The experimental water diffusion increases with increasing hydration level toward the value for bulk water . 
%As expected, water translational diffusion increase in the diffusion is observed with the increased temperature and hydration level (Figs.~\ref{fig:permeability} H and~\ref{fig:diffusionSI} B). 
Simulation results are close to the experimental values with low hydration levels, but increase to the values of approximately two times higher than experimental bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. The water diffusion parallel to membrane surface increased with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Fig.~\ref{fig:diffusionSI} in the supplementary information. In order to estimate the diffusion anistoropy of water in multilamellar membrane system, we translated the water permeation coefficients through membranes to diffusion coefficients through multilamellar stack using the Tanner equation~\cite{tanner78,wasterby02}. The resulting diffusion coefficients through the membranes are approximately five orders of magnitude slower than along the membrane (Figs.~\ref{fig:permeability} F and G), which is at the upper limits of the anisotropic estimated from the experimental data~\cite{nitsche19}. Relatively high anisotropy is understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} F and G.
This follows from decreasing water permeability with membrane packing (Figs \ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Fig. \ref{fig:diffusionSI} A and C).

>>>>

\begin{figure}[tb]
    \centering
    \includegraphics[width=180mm]{Figures/permeation2.pdf}
    \caption{A 
    {\bf B-E} Water permeation through membranes analyzed from the databank as a function of temperature, thickness, area per lipid, and acyl chain order. Values from simulations with non-zero permeation values are shown with blue dots. Averages over fixed range of x-axis values are shown with black dots. Insert in B) shows the Arrhenius plot of permeation ($\ln$ (P) vs. 1/T) that gives 17\,$\pm$\,4\,k$_b$T for the average activation energy for water permeation through lipid bilayer. Inserts in C) and D) show the region where the dependence could be considered approximately linear.
    {\bf F} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.
    {\bf G-H} Diffusion anistoropy of water as a function of thickness and area per lipid. }
    \label{fig:permeability}
\end{figure}

To analyze the anisotropic diffusion of water in membrane environment, we also calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank. The results at different hydration levels are shown in Fig.~\ref{fig:permeability} F together with the experimental data~\cite{rudakova04}. 
%The experimental water diffusion increases with increasing hydration level toward the value for bulk water . 
%As expected, water translational diffusion increase in the diffusion is observed with the increased temperature and hydration level (Figs.~\ref{fig:permeability} H and~\ref{fig:diffusionSI} B). 
Simulation results are close to the experimental values with low hydration levels, but increase to the values of approximately two times higher than experimental bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. The water diffusion parallel to membrane surface increased with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Fig.~\ref{fig:diffusionSI} in the supplementary information. In order to estimate the diffusion anistoropy of water in multilamellar membrane system, we translated the water permeation coefficients through membranes to diffusion coefficients through multilamellar stack using the Tanner equation~\cite{tanner78,wasterby02}. The resulting diffusion coefficients through the membranes are approximately five orders of magnitude slower than along the membrane (Figs.~\ref{fig:permeability} G and H), which is at the upper limits of the anisotropic estimated from the experimental data~\cite{nitsche19}. Relatively high anisotropy is understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs \ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Fig. \ref{fig:diffusionSI} A and C).
